---
title: "Getting started with mixology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mixology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mixology)
library(dplyr)
```

## Overview

The `mixology` package provides two sentiment lexicons developed for opinion
mining of Covid-19 Twitter data, plus helper functions for tokenisation,
negation handling, and weighted scoring.

| Resource | Terms | Positive | Negative | Ambiguous |
|---|---|---|---|---|
| `mixology_covid_lexicon` | 4,166 | 1,953 (46.9%) | 1,924 (46.2%) | 289 (6.9%) |
| `mixology_lexicon` | 16,528 | 5,716 (34.6%) | 9,655 (58.4%) | 1,157 (7.0%) |

## Basic usage

### 1. Access the lexicons

```{r}
lex_covid <- get_lexicon("covid")
head(lex_covid)

# Inspect high-weight terms (frequent in the reference corpus)
lex_covid |>
  filter(weight > 2.5) |>
  arrange(desc(weight))
```

### 2. Tokenise tweets

```{r}
tweets <- c(
  "The lockdown is terrible and completely unjustified",
  "The vaccine rollout has been excellent, great progress",
  "I am not sure about the booster, it makes me nervous"
)

toks <- mixology_tokenize(tweets)
head(toks)
```

### 3. Basic sentiment scoring (unweighted, no negation)

```{r}
results <- mixology_sentiment(tweets, lexicon = "covid", weighted = FALSE)
results
```

### 4. Weighted scoring

Each token match is multiplied by the lexicon `weight` column
(log-normalised corpus frequency, range 0.5–3.0). High-frequency
domain terms like *lockdown* or *vaccine* contribute more to the score.

```{r}
results_w <- mixology_sentiment(tweets, lexicon = "covid", weighted = TRUE)
results_w
```

### 5. Negation-aware scoring

```{r}
results_neg <- mixology_sentiment(
  tweets,
  lexicon          = "covid",
  weighted         = TRUE,
  handle_negation  = TRUE,
  negation_window  = 3
)
results_neg
```

Notice that "not sure" in tweet 3 reverses the ambiguous polarity of *sure*.

### 6. Apply to a data frame

```{r eval=FALSE}
library(dplyr)

corpus <- read.csv("your_corpus.csv")  # must have a 'text' column

scores <- mixology_sentiment(
  corpus$text,
  lexicon         = "covid",
  weighted        = TRUE,
  handle_negation = TRUE
)

corpus <- bind_cols(corpus, scores |> select(-doc_id))
```

## Working with the stop words and negations

```{r}
# Inspect
head(stop_words_en)
negations_en

# Add custom stop words (e.g. Twitter-specific noise)
toks_custom <- mixology_tokenize(
  tweets,
  custom_stopwords = c("rt", "via", "amp")
)
```

## Comparing the two lexicons

```{r}
r_covid <- mixology_sentiment(tweets, lexicon = "covid")
r_full  <- mixology_sentiment(tweets, lexicon = "full")

comparison <- bind_rows(
  mutate(r_covid, lexicon = "Covid"),
  mutate(r_full,  lexicon = "Full")
) |>
  select(doc_id, lexicon, coverage, score_net, polarity)

comparison
```

## Notes on the weight column

The `weight` column encodes the log-normalised frequency of each term in
the Mixology political measures sub-corpus (n = 4,371 tweets):

```
weight = 0.5 + (log(freq + 1) / log(max_freq + 1)) * 2.5
```

- Terms **unseen** in the reference corpus receive `weight = 0.5`.
- The **most frequent** term (*covid*, n = 2,015) receives `weight = 3.0`.
- The scale is continuous and can be used as-is or re-scaled.

If you want to weight by frequencies from **your own corpus**, you can
recompute the column:

```{r eval=FALSE}
library(dplyr)

# your_freq: a named numeric vector of token frequencies
lex <- get_lexicon("covid")
max_f <- max(your_freq)

lex <- lex |>
  mutate(
    freq_corpus = as.integer(your_freq[word]),
    freq_corpus = if_else(is.na(freq_corpus), 0L, freq_corpus),
    weight = 0.5 + (log(freq_corpus + 1) / log(max_f + 1)) * 2.5
  )
```
