---
title: "Getting started with mixology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mixology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mixology)
library(dplyr)
```

## Overview

The `mixology` package provides eight sentiment lexicons for comparative
opinion mining, two of which were developed specifically for Covid-19
crisis communication research. All lexicons share a common three-category
polarity scheme — **positive**, **negative**, **ambiguous** — making
cross-lexicon comparison straightforward.

| Short name | Full name | Terms |
|---|---|---|
| `inquirer` | General Inquirer | 4,206 |
| `subjectivity` | MPQA Subjectivity | 6,884 |
| `bing` | Bing Liu | 6,783 |
| `nrc` | NRC Emotion Lexicon | 6,456 |
| `afinn` | AFINN | 2,477 |
| `loughran` | Loughran-McDonald | 3,917 |
| `covid` | Mixology Covid Lexicon | 4,166 |
| `mixology` | Mixology Lexicon (merged) | 16,528 |

---

## 1. Access lexicons

```{r}
# List all available lexicons
mixology_lexicon_names()

# Retrieve any lexicon as a tibble
bing <- get_lexicon("bing")
head(bing)

# The Mixology lexicons also include weight and freq_corpus
covid <- get_lexicon("covid")
head(covid)

# High-weight terms (frequent in the reference corpus)
covid |> filter(weight > 2.5) |> arrange(desc(weight)) |> head(10)
```

---

## 2. Tokenise text

```{r}
tweets <- c(
  "The lockdown is terrible and completely unjustified",
  "The vaccine rollout has been excellent, great progress",
  "I am not sure about the booster at all"
)

toks <- mixology_tokenize(tweets)
head(toks)
```

Add custom stop words (e.g. Twitter noise):

```{r}
toks_custom <- mixology_tokenize(tweets,
  custom_stopwords = c("rt", "via", "amp"))
```

---

## 3. Sentiment scoring

### Basic (unweighted, no negation)

```{r}
mixology_sentiment(tweets, lexicon = "covid", weighted = FALSE)
```

### Weighted scoring

Matches are multiplied by the `weight` column — high-frequency domain terms
like *lockdown* or *vaccine* contribute more to the score.

```{r}
mixology_sentiment(tweets, lexicon = "covid", weighted = TRUE)
```

### Negation-aware scoring

```{r}
mixology_sentiment(tweets,
  lexicon         = "covid",
  weighted        = TRUE,
  handle_negation = TRUE,
  negation_window = 3
)
```

In tweet 3, *not sure* reverses the polarity of *sure* within the window.

### Applying to a data frame

```{r eval=FALSE}
corpus <- read.csv("your_corpus.csv")  # needs a 'text' column

scores <- mixology_sentiment(corpus$text, lexicon = "covid",
  weighted = TRUE, handle_negation = TRUE)

corpus <- bind_cols(corpus, scores |> select(-doc_id))
```

---

## 4. Compare all lexicons

### Corpus-level summary

```{r}
compare_lexicons(tweets)
```

The summary table shows, for each lexicon: number of terms, coverage,
percentage of documents classified positive / negative / ambiguous, and
mean net score. This is the main tool for comparative analysis.

### Document-level results (long format)

```{r}
compare_lexicons(tweets,
  lexicons = c("bing", "nrc", "covid", "mixology"),
  summary  = FALSE
)
```

### Visualise with ggplot2

```{r eval=FALSE}
library(ggplot2)

comp <- compare_lexicons(tweets)

ggplot(comp, aes(x = reorder(lexicon_label, pct_negative),
                 y = pct_negative)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "% tweets classified negative by lexicon",
       x = NULL, y = "%") +
  theme_minimal()
```

---

## 5. Coverage diagnostic

Quick check of how many corpus tokens each lexicon matches — before
running the full analysis:

```{r}
lexicon_coverage(tweets)
```

---

## 6. Identify cross-lexicon conflicts

Find terms that carry different polarities across lexicons:

```{r}
# Conflicts between the Covid lexicon and Bing
conflicts <- lexicon_conflicts(c("covid", "bing"))
head(conflicts, 15)
```

```{r eval=FALSE}
# All eight lexicons — terms with 3+ distinct polarities
lexicon_conflicts() |> filter(n_distinct >= 3)
```

These conflicts are informative: they reveal where domain-specific
annotation (as in the Covid lexicon) diverges from general-purpose
resources, and can guide further manual review.

---

## 7. Recompute weights from your own corpus

The default weights are derived from the Mixology political measures
sub-corpus (n = 4,371 tweets, Western Europe, December 2021). To adapt
them to a different corpus:

```{r eval=FALSE}
# your_freq: a named integer vector (names = tokens, values = counts)
# e.g. built with table() on your tokenised corpus

max_f <- max(your_freq)

lex <- get_lexicon("covid") |>
  mutate(
    freq_corpus = coalesce(your_freq[word], 0L),
    weight      = 0.5 + (log(freq_corpus + 1) / log(max_f + 1)) * 2.5
  )
```

---

## Notes on design choices

**Ambiguity over neutrality.** Terms that can be positive or negative
depending on context (e.g. *tested positive*, *government*, *restrictions*)
are labelled `"ambiguous"` rather than excluded or forced into a binary.
This is especially important for Covid-related discourse.

**Near-balanced Covid lexicon.** Unlike general-purpose dictionaries —
which typically contain a large surplus of negative terms — the Mixology
Covid Lexicon has a near-equal distribution (46.9% positive, 46.2%
negative). This reduces the systematic negative bias observed in general
lexicons when applied to domain-specific corpora.

**Unigram baseline.** All functions operate on unigrams by default.
Negation handling adds a sliding-window approximation. Multi-word
expressions and idiomatic usage are not addressed.
